---
title: "ASSIGNMENT_8_0733_MDTS4214"
author: "ROUNAK SENGUPTA"
date: "2026-03-30"
output: word_document
---

# Problem: Logistic Regression with Binary Response

We use the `Weekly` data from the `ISLR` package to demonstrate logistic regression for a binary response variable.

## (a) Train-test split

We use observations from **1990 to 2008** as the training data and observations from **2009 to 2010** as the test data.

```{r}
library(ISLR)
data("Weekly")

train_data <- subset(Weekly, Year >= 1990 & Year <= 2008)
test_data  <- subset(Weekly, Year >= 2009 & Year <= 2010)

dim(train_data)
dim(test_data)
```

The response variable is `Direction`, and the predictors are `Lag1`, `Lag2`, `Lag3`, `Lag4`, `Lag5`, and `Volume`.

# (b) Logistic regression with five lag variables and Volume

We now fit a logistic regression model on the training data.

```{r}
glm_full <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume,
                data = train_data,
                family = binomial)

summary(glm_full)
```

## Significance of predictors

The p-values of the predictors are:

```{r}
full_summary <- summary(glm_full)
coef_table <- full_summary$coefficients
coef_table
```

We now identify whether any predictor is significant at the 5% level.

```{r}
pvals <- coef_table[-1, 4]   # exclude intercept
significant_predictors <- names(pvals[pvals < 0.05])
significant_predictors
```

If no predictor is significant at the 5% level, then we find the predictor with the smallest p-value.

```{r}
least_p_name <- names(which.min(pvals))
least_p_value <- min(pvals)

least_p_name
least_p_value
```

## Interpretation of the coefficient with the least p-value

The corresponding coefficient is:

```{r}
least_p_coef <- coef(glm_full)[least_p_name]
least_p_coef
```

Its odds ratio is:

```{r}
exp(least_p_coef)
```

### Interpretation

For a one-unit increase in `r least_p_name`, keeping all other predictors fixed, the **log-odds** of the market going **Up** changes by `r round(least_p_coef, 4)`.

Equivalently, the **odds** are multiplied by `r round(exp(least_p_coef), 4)`.

- If this value is greater than 1, the odds of market going **Up** increase.
- If this value is less than 1, the odds of market going **Up** decrease.

# (c) Odds of market going "Up" at median predictor values

We compute the median of each predictor using the training data.

```{r}
med_vals <- data.frame(
  Lag1 = median(train_data$Lag1),
  Lag2 = median(train_data$Lag2),
  Lag3 = median(train_data$Lag3),
  Lag4 = median(train_data$Lag4),
  Lag5 = median(train_data$Lag5),
  Volume = median(train_data$Volume)
)

med_vals
```

Now we estimate the probability of the market going **Up** at these median predictor values.

```{r}
p_up_median <- predict(glm_full, newdata = med_vals, type = "response")
p_up_median
```

The estimated **odds** are:

```{r}
odds_up_median <- p_up_median / (1 - p_up_median)
odds_up_median
```

Thus, at the median values of the predictors, the estimated odds that the market goes **Up** are `r round(odds_up_median, 4)`.

# (e) Optimal cut-point on the test data

We now use the fitted model to obtain predicted probabilities for the test data.

```{r}
test_prob_full <- predict(glm_full, newdata = test_data, type = "response")
head(test_prob_full)
```

We define a function to compute:

- True Positive Rate (TPR)
- False Positive Rate (FPR)
- Objective: `TPR * (1 - FPR)`

for a given cut-point.

```{r}
evaluate_cutpoint <- function(actual, prob, cutoff) {
  pred <- ifelse(prob > cutoff, "Up", "Down")
  pred <- factor(pred, levels = c("Down", "Up"))
  actual <- factor(actual, levels = c("Down", "Up"))

  cm <- table(Predicted = pred, Actual = actual)

  # make sure all cells exist
  if (!("Down" %in% rownames(cm))) {
    cm <- rbind(cm, Down = c(0, 0))
  }
  if (!("Up" %in% rownames(cm))) {
    cm <- rbind(cm, Up = c(0, 0))
  }
  if (!("Down" %in% colnames(cm))) {
    cm <- cbind(cm, Down = c(0, 0))
  }
  if (!("Up" %in% colnames(cm))) {
    cm <- cbind(cm, Up = c(0, 0))
  }

  cm <- cm[c("Down", "Up"), c("Down", "Up")]

  TN <- cm["Down", "Down"]
  FN <- cm["Down", "Up"]
  FP <- cm["Up", "Down"]
  TP <- cm["Up", "Up"]

  TPR <- ifelse((TP + FN) == 0, 0, TP / (TP + FN))
  FPR <- ifelse((FP + TN) == 0, 0, FP / (FP + TN))
  objective <- TPR * (1 - FPR)
  error_rate <- (FP + FN) / sum(cm)

  list(
    cutoff = cutoff,
    confusion_matrix = cm,
    TPR = TPR,
    FPR = FPR,
    objective = objective,
    error_rate = error_rate
  )
}
```

Now we search over possible cut-points.

```{r}
cut_grid <- sort(unique(test_prob_full))
results_full <- lapply(cut_grid, function(cut) {
  evaluate_cutpoint(test_data$Direction, test_prob_full, cut)
})

objective_values_full <- sapply(results_full, function(x) x$objective)
best_index_full <- which.max(objective_values_full)
best_result_full <- results_full[[best_index_full]]

best_result_full$cutoff
best_result_full$TPR
best_result_full$FPR
best_result_full$objective
```

## Confusion matrix for the optimal cut-point

```{r}
best_result_full$confusion_matrix
```

## Test error rate

```{r}
best_result_full$error_rate
```

So, using the optimal cut-point, the test error rate is `r round(best_result_full$error_rate, 4)`.

# (f) Logistic regression using only Lag1 and Lag2

Now we fit another model using only `Lag1` and `Lag2`.

```{r}
glm_lag12 <- glm(Direction ~ Lag1 + Lag2,
                 data = train_data,
                 family = binomial)

summary(glm_lag12)
```

We obtain predicted probabilities for the test data.

```{r}
test_prob_lag12 <- predict(glm_lag12, newdata = test_data, type = "response")
head(test_prob_lag12)
```

Again, we find the optimal cut-point by maximizing `TPR * (1 - FPR)`.

```{r}
cut_grid_lag12 <- sort(unique(test_prob_lag12))
results_lag12 <- lapply(cut_grid_lag12, function(cut) {
  evaluate_cutpoint(test_data$Direction, test_prob_lag12, cut)
})

objective_values_lag12 <- sapply(results_lag12, function(x) x$objective)
best_index_lag12 <- which.max(objective_values_lag12)
best_result_lag12 <- results_lag12[[best_index_lag12]]

best_result_lag12$cutoff
best_result_lag12$TPR
best_result_lag12$FPR
best_result_lag12$objective
```

## Confusion matrix for the Lag1-Lag2 model

```{r}
best_result_lag12$confusion_matrix
```

## Test error rate for the Lag1-Lag2 model

```{r}
best_result_lag12$error_rate
```

# Comparison of test error rates

```{r}
full_error <- best_result_full$error_rate
lag12_error <- best_result_lag12$error_rate

comparison <- data.frame(
  Model = c("Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume", "Lag1 + Lag2"),
  Test_Error_Rate = c(full_error, lag12_error)
)

comparison
```

## Conclusion

```{r}
if (lag12_error < full_error) {
  cat("The model with only Lag1 and Lag2 gives a lower test error rate.")
} else if (lag12_error > full_error) {
  cat("The model with all six predictors gives a lower test error rate.")
} else {
  cat("Both models give the same test error rate.")
}
```

 